A Bayesian approach to uncertainty in word embedding bias estimation

### 2023-10-19 paper/CLsubmission/bayesBiasCL.tex

\usepackage{amsmath,amssymb}
\usepackage{float}
\usepackage{nicefrac}

\usepackage{graphicx}

@@

It has been suggested\footnote{See for instance \citep{Bolukbasi2016man, Caliskan2017semanticsBiases, Gonen2019lipstick, Lauscher2019multidimensional, Garg2018years, Manzini2019blackToCriminal}.}
that language models can learn implicit biases that reflect harmful
stereotypical thinking---for example, the (vector corresponding to the)
word \textit{she} might be much closer in the vector space to the word
\textit{cooking} than the word \textit{he}. Such phenomena are
undesirable at least in some downstream tasks, such as web search,
recommendations, and so on. To investigate such issues, several measures
of bias in word embeddings have been formulated and applied. Our goal is
to use two prominent examples of such measures to argue that this
approach oversimplifies the situation and to develop a Bayesian
alternative.

@@

Most of the problems that we point out generalize to any existing
approach that focuses on chasing a single numeric metric of bias. (1)
They treat the results of pre-averaging as raw data in statistical
significance tests, which in this context is bound to overestimate
significance. We show similar results can easily be obtained when
sampling from null models with no bias. (2) The word list sizes and
sample sizes used in the studies are usually small,\footnote{Depending
  on a list for \citep{Caliskan2017semanticsBiases} the range for protected words is between 13 and
  100, and for attributes between 16 and 25; for \citep{Manzini2019blackToCriminal} the range for
  protected words is between 14 and 18, and for attributes between 11
  and 25.} (3) Many studies do not use any control predicates, such as
random neutral words or neutral human predicates for comparison.

On the constructive side, we develop and deploy our method, and the
results are, roughly, as follows. (A) Posterior density intervals are
fairly wide and the average differences between associated, different
and neutral human predicates, are not very large. (B) A preliminary
inspection suggests that the desirability of changes obtained by the
usual debiasing methods is debatable.

In Section \ref{sec:two} we describe the two key measures discussed in
this paper, \textsf{WEAT} and \textsf{MAC}, explaining how they are
calculated and how they are supposed to work. In Section
\ref{sec:challenges} we first argue in Subsection
\ref{subsec:interpretability}, that it is far from clear how results
given in terms of \textsf{WEAT} or \textsf{MAC} are to be interpreted.
Second, in Subsection \ref{subsec:problems} we explain the statistical
problems that arise when one uses pre-averaged data in such contexts, as
these measures do. In Section \ref{sec:bayesian} we explain the
alternative Bayesian approach that we propose. In Section
\ref{sec:results} we elaborate on the results that it leads to,
including a somewhat skeptical view of the efficiency of debiasing
methods, discussed in Subsection \ref{subsec:rethinking}. Finally, in
Section \ref{sec:related} we spend some time placing our results in the
ongoing discussions.\footnote{\textbf{Disclaimer:} throughout the paper
  we will be mentioning and using word lists and stereotypes we did not
  formulate, which does not mean we condone any judgment made therein or
  underlying a given word selection. For instance, the Gender dataset
  does not recognize non-binary categories, and yet we use it without
  claiming that such categories should be ignored.}$^{, }$ \footnote{A few more philosophical comments on the enterprise of reflecting on bias in language models can be found in Appendix \ref{appendix:philosophical}.}

@@

One of the first measures of bias has been developed in \citep{Bolukbasi2016man}. The
general idea is that a certain topic is associated with a vector of real
numbers (the topic ``direction''), and the bias of a word is
investigated by considering the projection of its corresponding vector
on this direction. For instance, in \citep{Bolukbasi2016man}, the gender direction
\(\mathsf{gd}\) is obtained by taking the differences of the vectors
corresponding to ten different gendered pairs (such as
\(\overrightarrow{she} - \overrightarrow{he}\) or
\(\overrightarrow{girl} - \overrightarrow{boy}\)), and then identifying
their principal component.\footnote{Roughly, the principal component is
  the vector obtained by projecting the data points on their linear
  combination in a way that maximizes the variance of the projections.}
The gender bias of a word \(W\) is then understood as \(W\)'s projection
on the gender direction: \(\vec{W} \cdot GD\) (which, after normalizing
by dividing by \(\lVert W \rVert \,\lVert \mathsf{GD} \rVert\), is the
same as cosine similarity). Given a list \(\mathbf{N}\) of supposedly gender
neutral words,\footnote{We follow the methodology used in the debate in
  assuming that there is a class of words identified as more or less
  neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose
  average similarity to the gender direction (or other protected words)
  is around 0. See our list in Appendix \ref{app:custom} and a brief
  discussion in Subsection \ref{subsec:interpretability}.} and the
gender direction \(\mathsf{GD}\), the direct gender bias is defined as
the average cosine similarity of the words in \(\mathbf{N}\) from \(\mathsf{GD}\)
(\(c\) is a parameter determining how strict we want to be):
\begin{align*}
\mathsf{directBias_c(\mathbf{N},GD)} & = \frac{\sum_{W\in \mathbb{N}}\vert \mathsf{cos}(\vec{W},\mathsf{GD})\vert^c}{\vert \mathbf{N} \vert }
\end{align*} \normalsize

The use of projections in bias estimation has been criticized for
instance in \citep{Gonen2019lipstick}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, it is only one possible manifestation
of it, and reducing the cosine similarity to such a projection may not
be sufficient to eliminate bias. For instance, ``math'' and ``delicate''
might be equally similar to a pair of opposed explicitly gendered words
(\emph{she}, \emph{he}), while being closer to quite different
stereotypical attribute words (such as \emph{scientific} or
\emph{caring}). Further, it is observed in \citep{Gonen2019lipstick} that most word pairs
retain similarity under debiasing meant to minimize projection-based
bias.\footnote{In \citep{Bolukbasi2016man} another method that involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  \citep{Nissim2020fair}.}

@@

\textsf{WEAT} has been developed to investigate biases corresponding to
a pair of supposedly opposing stereotypes, so the question arises as to
how to generalize the measure to contexts in which biases with respect
to more than two stereotypical groups are to be measured. Such a
generalization can be found in \citep{Manzini2019blackToCriminal}. The authors introduce Mean
Average Cosine distance (\textsf{MAC}) as a measure of bias. Let
\(T = \{t_1, \dots, t_k\}\) be a set of protected words, and let each
\(A_j\in \mathcal{A}\) be a set of attributes stereotypically associated
with a protected word where \(\mathcal{A}\). For instance, when biases
related to religion are to be investigated, they use a dataset of the
format illustrated in Table \ref{tab:religionOriginal}. The measure is
defined as follows:

@@

\noindent That is, for each protected word \(t\in T\), and each
attribute set \(A_j\), they first take the mean of distances for this
protected word and all attributes in a given attribute class, and then
take the mean of thus obtained means for all the protected words and all
the protected classes.\footnote{The authors' code is available through
  their github repository at
  \url{https://github.com/TManzini/DebiasMulticlassWordEmbedding}.}

@@

Table \ref{tab:religionOriginal2} contains an example of \textsf{MAC}
scores (and \(p\) values, we explain how these are obtained in
Subsection \ref{subsec:problems}) before and after deploying two
debiasing methods to the Reddit embedding, where the score is calculated
using the Religion word lists from \citep{Manzini2019blackToCriminal}. For our purpose the details
of the debiasing method are not important: what matters is that
\textsf{MAC} is used in the evaluation of these methods.

@@

The first question we should ask is whether the initial \textsf{MAC}
values lower than 1 indeed are indicative of the presence of bias?
Thinking abstractly, 1 is the ideal distance for unrelated words. But in
fact, there is some variation in distances, which might lead to
non-biased lists also having \textsf{MAC} scores smaller than 1. How
much smaller? What may attract attention is the fact that the value of
cosine distance in ``Biased'' category is already quite high (i.e.~close
to 1) even before debiasing. High cosine distance indicates low cosine
similarity between values. One could think that the average cosine
similarity equal to approximately 0.141 is not large enough to claim the
presence of a bias to start with. The authors, though, still aim to
mitigate it by making the distances involved in the \textsf{MAC}
calculations even larger. The question is, on what basis is this small
similarity still considered as proof of the presence of bias, and
whether these small changes are meaningful.

@@

In Figures (\ref{fig:empirical0}-\ref{fig:empirical3}) we look at the
empirical distributions, while paying attention to such divisions. The
horizontal lines represent the values of \(1 - \mathsf{MAC}\) (that is,
we now talk in terms of cosine similarity rather than cosine distance)
that the authors considered indicative of bias for stereotypes
corresponding to given word lists. For instance, in religion,
\textsf{MAC} was .859, which was considered a sign of bias, so we plot
\(0\pm (1-.859)\approx .14\) lines around similarity = 0 (that is,
distance = 1). Notice that most distributions are quite wide, and the
proportions of even neutral or human neutral words with similarities
higher than the value of \(1 - \mathsf{MAC}\) deserving debiasing
according to the authors are quite high.

@@

\caption{Empirical distributions of cosine similarities  for the Religion word list  used in  the original paper. }

@@

\caption{Empirical distributions of cosine similarities for the Gender word list used in  the original paper.  }

@@

\caption{Empirical distributions of cosine similarities  for the Race word list used in  the original paper.  }

@@

As transparent in Figures \ref{fig:muslim} and \ref{fig:priest}, for the
protected word \textsf{muslim}, the most similar attributes tend to be
the ones associated with it stereotypically, but then words associated
with other stereotypes come closer than neutral or human predicates. For
the protected word \textsf{priest}, the situation is even less as
expected: the nearest attributes are human attributes, and all there
seems to be no clear pattern to the distances to other attributes.

@@

In Figure \ref{fig:caliskanCalc} we illustrate one iteration of the
procedure. We draw one such sample of size \(16^2\). Then we actually
list all possible ways to split the 16 words in two equal sets (each
such a split is one bootstrapped sample) and for each of them we
calculate the \textsf{s} values and \textsf{WEAT}. What are the
resulting distributions of \(\textsf{s}\) scores and what \(p\)-values
do they lead to? What are the resulting effect sizes for each
bootstrapped sample, and how often can we get an effect size as large as
the ones reported in the original paper?

In the bootstrapped samples we would rather expect low \textsf{s} values
and low \textsf{WEAT}: after all, these are just random permutations of
random distances all of which are drawn from the same null distribution.
Let's take a look at one such a bootstrapped sample.

On purpose, we picked a rather unusual one: the observed test statistic
is 0.39 and 1.27. The bootstrapped distributions of the test statistics
and effect sizes are illustrated in Figure \ref{fig:caliskanCalc},
together with this particular example. Quite notably both (two-sided)
\(p\) values for our example are rather low (Figure
\ref{fig:caliskanCalc}). These facts might suggest that we ended up
with a situation where ``bias'' is present (albeit, due to random
noise). The reason why we picked it is that it is an example of a word
list that ends up with relatively low \(p\)-value and a relatively
unusual effect size, but nevertheless, its closer inspection shows that
even for a word list with such properties there is no clear reason to
think that the bias is present.

@@

At this point, we might think that we just stumbled into a bootstrapped
sample that randomly happened to display strong bias. We decide to
double-check this by visual inspection expecting exactly this: a strong,
clearly visible bias (Figure \ref{fig:caliskanDistances}).

@@

\noindent In fact, while there might be some outliers here and there,
saying that a clear bias on which one group is systematically closer to
\(A\)s than another is definitely a stretch. What happened?

In the calculations of \textsf{WEAT} means are taken twice. The
\textsf{s}-values themselves are means and then means of
\textsf{s}-values are compared between groups. Statistical troubles
start when we run statistical tests on sets of means, for at least two
reasons.

@@

Let us think again about the question that we are ultimately interested
in. Are the \(X\) terms systematically closer to (further from) the
\(A\) attributes (\(B\) attributes) than the \(Y\) words? But now let's
use the raw data points to try to answer these questions.

To start with, let us run two quick \(t\)-tests to gauge what the raw
data illustrated in Figure \ref{fig:caliskanDistances} tell us. First,
distances to \(A\) attributes for \(X\) words and \(Y\) words. Well, the
result is---strictly speaking---statistically significant. The
\(p\)-value is \(0.02\) (more than ten times higher than the
\(p\)-valued obtained by the bootstrapping procedure. So the sample is
in some sense unusual. But the 95\% confidence interval for the
difference in means is \([.0052, .061]\), clearly nothing that a reader
would expect given that the calculated effect size seemed quite large.
How about the distances to the \(B\) attributes? Here the \(p\)-value is
\(.22\) and the 95\% confidence interval is \([-0.03, .009]\), even less
of a reason to think a bias is present.

The difficulties are exacerbated by the fact that statistical tests are
based on bootstrapping from a relatively small data sets, which is quite
likely to underestimate the population variance. To make our point
clear, let us avoid bootstrapping and work with the null generative
model with \(\mathsf{Norm}(0,.08)\) for both word groups. We keep the
sizes the same: we have eight protected words in each group, sixteen in
total, and for each we randomly draw 8 distances from hypothetical \(A\)
attributes, and \(8\) distances from hypothetical \(B\) attributes.
Calculate the test statistic and effect size the way \citep{Caliskan2017semanticsBiases} did. Do
this 10000 times, each time calculating \textsf{WEAT} and \textsf{s}
values, and look at what the distributions of these values are on the
assumption of the null model with realistic empirically motivated raw
data point standard deviation.

@@

What is the alternative? As we already emphasized: focusing on what the
real underlying question is and trying to answer it using a statistical
analysis of the raw data using meaningful control groups, to ensure
interpretability. Moreover, since the data sets are not too large and
since multiple evaluations are to be made, we will pursue this method
from the Bayesian perspective. Now we have to describe it.

@@

Bayesian data analysis takes prior probability distributions, a
mathematical model structure and the data, and returns the posterior
probability distributions over the parameters of interest, thus
capturing our uncertainty about their actual values. One important
difference between such a result and the result of classical
statistical analysis is that classical confidence intervals (CIs) have a
rather complicated and somewhat confusing interpretation, which has
little to do with the posterior probability distribution.\footnote{Here
  are a few usual problems. CIs are often mistakenly interpreted as
  providing the probability that a resulting confidence interval
  contains the true value of a parameter. CIs bring confusion also with
  regard to precision, it is a common mistake to interpret narrow
  intervals as the ones corresponding to more precise knowledge.
  Another fallacy is to associate CIs with likelihood and to state that
  values within a given interval are more probable than the ones outside
  it. The theory of confidence intervals does not support the above
  interpretations. CIs should be plainly interpreted as a result of
  a certain procedure (there are many ways to obtain CIs from a given set
  of data) that will in the long run contain the true value if the
  procedure is performed a fixed amount of times. For a nice survey and
  explanation of these misinterpretations, see \citep{Morey2015confidenceFallacy}. For a
  psychological study of the occurrence of such misinterpretations, see
  \citep{Hoekstra2014Misinterpretation}. In this study, 120 researchers and 442 students were asked to
  assess the truth value of six false statements involving different
  interpretations of a CI. Both researchers and students endorsed, on
  average, more than three of these statements.}

In fact, Bayesian highest posterior density intervals (HPDIs, the
narrowest intervals containing a certain ratio of the area under the
curve) and CIs end up being numerically the same only if the prior
distributions are completely uninformative. This illustrates that (1) classical analysis
is unable to incorporate non-trivial priors, and (2) is therefore more
susceptible to over-fitting, unless regularization (equivalent to a more
straightforward Bayesian approach) is used. In contrast with CIs, the
posterior distributions are easily interpretable and have direct
relevance to the question at hand. Moreover, Bayesian data analysis is
better at handling hierarchical models and small datasets, which is
exactly what we will we dealing with.

In standard Bayesian analysis, the first step is to understand the
data, think hard about the underlying process, and select potential
predictors and the outcome variable. The next step is to formulate a
mathematical description of the generative model of the relationships
between the predictors and the outcome variable. Prior distributions
must then be chosen for the parameters used in the model. Next, Bayesian
inference must be applied to find posterior distributions over the
possible parameter values. Finally, we need to check how well the
posterior predictions reflect the data with a posterior predictive
check.

In our analysis, the outcome variable is the cosine distances between the
protected words and attribute words. The predictor is a factor
determining whether a given attribute word is a neutral word, a human
predicate is stereotypically associated with the protected word, or
comes from a different stereotype connected with another protected word.
The idea is really straightforward: if bias is present in the embedding,
distances to associated attribute words should be systematically lower
than to other attribute words.

Furthermore, conceptually there are two levels of analysis in our
approach (see Figure \ref{fig:visDag}). On the one hand, we are
interested in the general question of whether related attributes are
systematically closer across the dataset. On the other hand, we are
interested in a more fine-grained picture of the role of the predictor
for particular protected words. Learning in hierarchical Bayesian models
involves using Bayesian inference to update the parameters of the model.
This update is based on the observed data, and estimates are made at
different levels of the data hierarchy. We use hierarchical Bayesian
models in which we simultaneously estimate parameters at the protected
word level and at the global level, assuming that all lower-level
parameters are drawn from global distributions. Such models can be
thought of as incorporating adaptive regularization, which avoids
overfitting and leads to improved estimates for unbalanced datasets (and
the datasets we need to use are unbalanced).

>>>>

\usepackage{amsmath,amssymb}
\usepackage{float}
\usepackage{nicefrac}
\usepackage{enumitem}
\usepackage{graphicx}

@@

It has been suggested  that language models can learn implicit biases that reflect harmful
stereotypical thinking (see for instance \citealt{Bolukbasi2016man, Caliskan2017semanticsBiases, Gonen2019lipstick, Lauscher2019multidimensional, Garg2018years, Manzini2019blackToCriminal}).
For example, the (vector corresponding to the)
word \textit{she} might be much closer in the vector space to the word
\textit{cooking} than the word \textit{he}. Such phenomena are
undesirable at least in some downstream tasks, such as web search,
recommendations, and so on. To investigate such issues, several measures
of bias in word embeddings have been formulated and applied. Our goal is
to use two prominent examples of such measures to argue that this
approach oversimplifies the situation and to develop a Bayesian
alternative.

@@

Most of the problems that we point out generalize to any existing
approach that focuses on chasing a single numeric metric of bias. (1)
They treat the results of pre-averaging as raw data in statistical
significance tests, which in this context is bound to overestimate
significance. We show similar results can easily be obtained when
sampling from null models with no bias. (2) The word list sizes and
sample sizes used in the studies are usually small,\footnote{Depending
  on a list for \citep{Caliskan2017semanticsBiases} the range for protected words is between 13 and
%  100, and for attributes between 16 and 25; for \citep{Manzini2019blackToCriminal} the range for
  protected words is between 14 and 18, and for attributes between 11
  and 25.} (3) Many studies do not use any control predicates, such as
random neutral words or neutral human predicates for comparison.

On the constructive side, we develop and deploy our method, and the
results are, roughly, as follows. (A) Posterior density intervals are
fairly wide and the average differences in cosine distances between stereotypically associated, stereotypically different,  
random neutral, and regular human-related predicates are not very large. (B) A preliminary inspection suggests that the desirability of changes obtained by the
usual debiasing methods is debatable.

In Section \ref{sec:two} we describe the two key measures discussed in
this paper, \textsf{WEAT} and \textsf{MAC}, explaining how they are
calculated and how they are supposed to work. In Section
\ref{sec:challenges} we first argue in Subsection
\ref{subsec:interpretability}, that it is far from clear how results
given in terms of \textsf{WEAT} or \textsf{MAC} are to be interpreted.
Second, in Subsection \ref{subsec:problems} we explain the statistical
problems that arise when one uses pre-averaged data in such contexts, as
these measures do. In Section \ref{sec:bayesian} we explain the
alternative Bayesian approach that we propose. In Section
\ref{sec:results} we elaborate on the results that it leads to,
including low  efficiency of debiasing methods, discussed in Subsection \ref{subsec:rethinking}. Finally, in
Section \ref{sec:related} we spend some time placing our results in the
ongoing discussions.\footnote{\textbf{Disclaimer:} throughout the paper
  we will be mentioning and using word lists and stereotypes we did not
  formulate, which does not mean we condone any judgment made therein or
  underlying a given word selection. For instance, the Gender dataset
  does not recognize non-binary categories, and yet we use it without
  claiming that such categories should be ignored.}$^{, }$ \footnote{A few more philosophical comments on the enterprise of reflecting on bias in language models can be found in Appendix \ref{appendix:philosophical}.}

@@

One of the first measures of bias has been developed in \citep{Bolukbasi2016man}. The
general idea is that a certain topic is associated with a vector of real
numbers (the topic ``direction''), and the bias of a word is
investigated by considering the projection of its corresponding vector
on this direction. For instance, in \citep{Bolukbasi2016man}, the gender direction
\(\mathsf{GD}\) is obtained by taking the differences of the vectors
corresponding to ten different gendered pairs (such as
\(\overrightarrow{she} - \overrightarrow{he}\) or
\(\overrightarrow{girl} - \overrightarrow{boy}\)), and then identifying
their principal component.\footnote{Roughly, the principal component is
  the vector obtained by projecting the data points on their linear
  combination in a way that maximizes the variance of the projections.}
The gender bias of a word \(W\) is then understood as \(W\)'s projection
on the gender direction: \(\vec{W} \cdot GD\) (which, after normalizing
by dividing by \(\lVert W \rVert \,\lVert \mathsf{GD} \rVert\), is the
same as cosine similarity). Given a list \(\mathbf{N}\) of supposedly gender
neutral words,\footnote{We follow the methodology used in the debate in
  assuming that there is a class of words identified as more or less
  neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose
  average similarity to the gender direction (or other protected words)
  is around 0. See our list in Appendix \ref{app:custom} and a brief
  discussion in Subsection \ref{subsec:interpretability}.} and the
gender direction \(\mathsf{GD}\), the direct gender bias is defined as
the average cosine similarity of the words in \(\mathbf{N}\) from \(\mathsf{GD}\)
(\(c\) is a parameter determining how strict we want to be):
\begin{align*}
\mathsf{directBias_c(\mathbf{N},GD)} & = \frac{\sum_{W\in \mathbb{N}}\vert \mathsf{cos}(\vec{W},\mathsf{GD})\vert^c}{\vert \mathbf{N} \vert }
\end{align*} \normalsize

The use of projections in bias estimation has been criticized for
instance in \citep{Gonen2019lipstick}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, it is only one possible manifestation
of it, and reducing the cosine similarity to such a projection may not
be sufficient to eliminate bias. For instance, \emph{math} and \emph{delicate} 
might be equally similar to a pair of opposed explicitly gendered words
(\emph{she}, \emph{he}), while being closer to quite different
stereotypical attribute words (such as \emph{scientific} or
\emph{caring}). Further, it is observed in \citep{Gonen2019lipstick} that most word pairs
retain similarity under debiasing meant to minimize projection-based
bias.\footnote{In \citep{Bolukbasi2016man} another method that involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  \citep{Nissim2020fair}.}

@@

\textsf{WEAT} has been developed to investigate biases corresponding to
a pair of supposedly opposing stereotypes, so the question arises as to
how to generalize the measure to contexts in which biases with respect
to more than two stereotypical groups are to be measured. Such a
generalization can be found in \citep{Manzini2019blackToCriminal}. The authors introduce Mean
Average Cosine distance (\textsf{MAC}) as a measure of bias. Let
\(T = \{t_1, \dots, t_k\}\) be a set of protected words, and let each
\(A_j\in \mathcal{A}\) be a set of attributes stereotypically associated
with a protected word. For instance, when biases
related to religion are to be investigated, they use a dataset of the
format illustrated in Table \ref{tab:religionOriginal}. The measure is
defined as follows:

@@

\noindent For each protected word \(t\in T\), and each
attribute set \(A_j\), they first take the mean of distances for this
protected word and all attributes in a given attribute class, and then
take the mean of thus obtained means for all the protected words and all
the protected classes.\footnote{The authors' code is available through
  their github repository at
  \url{https://github.com/TManzini/DebiasMulticlassWordEmbedding}.}

@@

Table \ref{tab:religionOriginal2} contains an example of \textsf{MAC}
scores (and \(p\) values, we explain how these are obtained in
Subsection \ref{subsec:problems}) before and after deploying two
debiasing methods to the Reddit embedding, where the score is calculated
using the Religion word lists from \citep{Manzini2019blackToCriminal}. The main element of the strategy is projection  onto (hard debiasing) or shifting vectors towards (soft debiasing) the bias subspace so that the biased information is no longer linearly accessible, and normalization of the vectors.\footnote{See the authors' GitHub code at \url{{https://github.com/TManzini/DebiasMulticlassWordEmbedding/blob/master/Debiasing/biasOps.py}}  for details.}  For our purpose, details of the debiasing methods are not important. What matters is that the authors use \textsf{MAC} in the evaluation of these methods.

@@

The first question we should ask is whether the initial \textsf{MAC}
values lower than 1 indeed are indicative of the presence of bias?
Thinking abstractly, 1 is the ideal distance for unrelated words. But in
fact, there is some variation in distances, which might lead to
non-biased lists also having \textsf{MAC} scores smaller than 1.
 What may attract attention is the fact that the value of
cosine distance in ``Biased'' category is already quite high (i.e.~close
to 1) even before debiasing. High cosine distance indicates low cosine
similarity between values. One could think that the average cosine
similarity equal to approximately 0.141 is not large enough to claim the
presence of a bias to start with. The authors, though, still aim to
mitigate it by making the distances involved in the \textsf{MAC}
calculations even larger. The question is, on what basis is this small
similarity still considered as proof of the presence of bias, and
whether these small changes are meaningful.

@@

In Figures (\ref{fig:empirical0}-\ref{fig:empirical3}) we look at the
empirical distributions, while paying attention to such divisions. The
horizontal lines represent the values of \(1 - \mathsf{MAC}\) (cosine similarity) that the authors considered indicative of bias for stereotypes corresponding to given word lists. For instance, in religion,
\textsf{MAC} was .859, which was considered a sign of bias, so we plot
\(0\pm (1-.859)\approx .14\) lines around similarity = 0 (that is,
distance = 1). Notice that most distributions are quite wide, and the
proportions of even neutral or human neutral words with similarities
higher than the value of \(1 - \mathsf{MAC}\) deserving debiasing
according to the authors are quite high.

@@

\caption{Empirical distributions of cosine similarities  for the Religion word list  used in  the original paper.  The
	horizontal lines represent the values of \(1 - \mathsf{MAC}\) (cosine similarity) that the authors considered indicative of bias for stereotypes corresponding to given word lists. }

@@

\caption{Empirical distributions of cosine similarities for the Gender word list used in  the original paper.  The
	horizontal lines represent the values of \(1 - \mathsf{MAC}\) (cosine similarity) that the authors considered indicative of bias for stereotypes corresponding to given word lists.}

@@

\caption{Empirical distributions of cosine similarities  for the Race word list used in  the original paper.  The
	horizontal lines represent the values of \(1 - \mathsf{MAC}\) (cosine similarity) that the authors considered indicative of bias for stereotypes corresponding to given word lists. }

@@

As transparent in Figures \ref{fig:muslim} and \ref{fig:priest}, for the
protected word \textsf{muslim}, the most similar attributes tend to be
the ones associated with it stereotypically, but then words associated
with other stereotypes come closer than neutral or human predicates. For
the protected word \textsf{priest}, the situation is even less as
expected: the nearest attributes are human attributes, and there
seems to be no clear pattern when it comes to the distances between attributes.

@@

In Figure \ref{fig:caliskanCalc} we illustrate one iteration of the
procedure. We draw one such sample of size \(16^2\). Then we actually
list all possible ways to split the 16 words in two equal sets (each
such a split is one bootstrapped sample) and for each of them we
calculate the \textsf{s} values and \textsf{WEAT}. What are the
resulting distributions of \(\textsf{s}\) scores,  what \(p\)-values and effect sizes
do they lead to?

In the bootstrapped samples we would rather expect low \textsf{s} values
and low \textsf{WEAT}: after all, these are just random permutations of
random distances all of which are drawn from the same null distribution. Let's take a look at one such a bootstrapped sample. For the sake of illustration,
 we picked a rather unusual one: the observed test statistic in this sample is 0.39 and 1.27.

 The bootstrapped distributions of the test statistics
and effect sizes across multiple samples are illustrated in Figure \ref{fig:caliskanCalc} (we marked the location of our particular example). 
Notably, both (two-sided) \(p\) values for our example are rather low.
This  might suggest that we ended up with a situation where ``bias'' is present (albeit, due to random
noise). After all the observed statistic is unusual enough for the \(p\) values to pass the traditional significance threshold.

@@

The reason why we picked the example that we did is that  while it leads to a relatively low \(p\)-value, and a relatively unusual effect size. Yet, a visualization of the sample reveals no interesting patterns (see Figure \ref{fig:caliskanDistances}), which strongly suggests that the way we calculated effect sizes and \(p\) values overestimates the impact of random noise, in line with our more theoretical comment, which will follow.

@@

\noindent In fact, while there might be some outliers here and there,
saying that a clear bias on which one group is systematically closer to
one word group  than another lacks support.  Crucially, in the calculations of \textsf{WEAT} means are taken twice. The \textsf{s}-values themselves are means and then means of \textsf{s}-values are compared between groups. Statistical troubles
start when we run statistical tests on sets of means, for at least two
reasons.

@@

Let us think again about the question that we are ultimately interested
in.  Are the \(X\) terms systematically closer to or further from the
\(A\)  or \(B\) attributes than the \(Y\) words?  But this time, instead of pre-averaging, let's use the raw data points to answer the question. To start with, let us run two quick \(t\)-tests to gauge what the raw
data illustrated in Figure \ref{fig:caliskanDistances} tell us. First,
distances to \(A\) attributes for \(X\) words and \(Y\) words. The
result is statistically significant. The
\(p\)-value is \(0.02\) (more than ten times higher than the
\(p\)-valued obtained by the bootstrapping procedure). So the sample is
in some sense unusual. But the 95\% confidence interval for the
difference in means is \([.0052, .061]\), which suggests values that are much smaller  than what a reader  would expect having read that the calculated effect size is quite large. Let us inspect the distances to the \(B\) attributes. Here the \(p\)-value is
\(.22\) and the 95\% confidence interval is \([-0.03, .009]\), even less
of a reason to think a bias is present.

Another difficulty is that these statistical tests are
based on bootstrapping from a relatively small data sets, which is quite
likely to underestimate the population variance. To make our point
clear, let us avoid bootstrapping and work with the null generative
model with \(\mathsf{Norm}(0,.08)\) for both word groups. We keep the
sizes the same: we have eight protected words in each group, sixteen in
total, and for each we randomly draw 8 distances from hypothetical \(A\)
attributes, and \(8\) distances from hypothetical \(B\) attributes.
Calculate the test statistic and effect size the way \citep{Caliskan2017semanticsBiases} did. Do
this 10000 times, each time calculating \textsf{WEAT} and \textsf{s}
values, and look at what the distributions of these values are on the
assumption of the null model with realistic empirically motivated raw
data point standard deviation (Figure \ref{fig:ourDistances}).

@@

As an alternative, we will propose focusing on what the
real underlying question is and trying to answer it using a statistical
analysis of the raw data using meaningful control groups, to ensure
interpretability. Moreover, since the data sets are not too large and
since multiple evaluations are to be made, we will pursue this method
from the Bayesian perspective.

@@

Bayesian data analysis takes prior probability distributions, a
mathematical model structure and the data, and returns the posterior
probability distributions over the parameters of interest, thus
capturing our uncertainty about their actual values. One important
difference between such a result and the result of classical
statistical analysis is that classical confidence intervals (CIs) have a
rather complicated and somewhat confusing interpretation not directly related to the posterior probability distribution.\footnote{Here
  are a few usual problems. CIs are often mistakenly interpreted as
  providing the probability that a resulting confidence interval
  contains the true value of a parameter. CIs bring confusion also with
  regard to precision, it is a common mistake to interpret narrow
  intervals as the ones corresponding to more precise knowledge.
  Another fallacy is to associate CIs with likelihood and to state that
  values within a given interval are more probable than the ones outside
  it. The theory of confidence intervals does not support the above
  interpretations. CIs should be plainly interpreted as a result of
  a certain procedure (there are many ways to obtain CIs from a given set
  of data) that will in the long run contain the true value if the
  procedure is performed a fixed amount of times. For a nice survey and
  explanation of these misinterpretations, see \citep{Morey2015confidenceFallacy}. For a
  psychological study of the occurrence of such misinterpretations, see
  \citep{Hoekstra2014Misinterpretation}. In this study, 120 researchers and 442 students were asked to
  assess the truth value of six false statements involving different
  interpretations of a CI. Both researchers and students endorsed, on
  average, more than three of these statements.}

In fact, Bayesian highest posterior density intervals\footnote{HPDIs are  the
narrowest intervals containing a certain ratio of the area under the
curve.} and CIs end up being numerically the same only if the prior
distributions are completely uninformative. This illustrates that (1) classical analysis
is unable to incorporate non-trivial priors, and (2) is therefore more
susceptible to over-fitting, unless regularization (equivalent to a more
straightforward Bayesian approach) is used. In contrast with CIs, the
posterior distributions are easily interpretable and have direct
relevance to the question at hand. Moreover, Bayesian data analysis is
better at handling hierarchical models and small datasets, which is
exactly what we will we dealing with.

In standard Bayesian analysis, the first step is to understand the
data, think hard about the underlying process, and select potential
predictors and the outcome variable. The next step is to formulate a
mathematical description of the generative model of the relationships
between the predictors and the outcome variable. Prior distributions
must then be chosen for the parameters used in the model. Next, Bayesian
inference must be applied to find posterior distributions over the
possible parameter values. Finally, we need to check how well the
posterior predictions reflect the data with a posterior predictive
check. This will be our evaluation of the methods performance, discussed in Subsection \ref{posterior-predictive-check}, and a more complete collection of such results is available in the appendix.

In our analysis, the outcome variable is the cosine distances between the
protected words and attribute words. The predictor is a factor related to an attribute word, and has  four levels:
\begin{itemize}[noitemsep]
	\item a given attribute word is stereotypically associated with the protected word,
	\item it comes from a different stereotype connected with another protected word,

	\item it is a neutral word,

	\item it is a human-related predicate.

\end{itemize} 
\noindent The idea is that if bias is present in the embedding,
distances to associated attribute words should be systematically lower
than to other attribute words.

Furthermore, conceptually there are two levels of analysis in our
approach (see Figure \ref{fig:visDag}). On the one hand, we are
interested in the general question of whether related attributes are
systematically closer across the dataset. On the other hand, we are
interested in a more fine-grained picture of the role of the predictor
for particular protected words. Learning in hierarchical Bayesian models
involves using Bayesian inference to update the parameters of the model.
This update is based on the observed data, and estimates are made at
different levels of the data hierarchy. We use hierarchical Bayesian
models in which we simultaneously estimate parameters at the protected
word level and at the global level, assuming that all lower-level
parameters are drawn from global distributions. Such models can be
thought of as incorporating adaptive regularization, which avoids
overfitting and leads to improved estimates for unbalanced datasets.
